Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-10-22 paper-new/discussion.tex
some fixes to paper based on reviews
paper revisions

% \eric{I moved this from the intro. We may need to revise it.}

@@

The methods of Abstracting Gradual Typing \cite{garcia-clark-tanter2016} and the
formal tools of the Gradualizer \cite{cimini-siek2016} provide a way to derive
from a statically-typed language a language that satisfies graduality by
construction. The main downside to these approaches lies in their inflexibility:
since the process in entirely mechanical, the language designer must adhere to
the predefined framework.  Many gradually typed languages do not fit into either
framework, e.g., Typed Racket \cite{tobin-hochstadt06, tobin-hochstadt08} and
the semantics produced is not always the desired one.
%
Furthermore, while these frameworks do prove graduality of the resulting
languages, they do not show the correctness of the equational theory, which is
equally important to sound gradual typing.

@@

% New and Licata \cite{new-licata18} developed an axiomatic account of
% the graduality relation on a call-by-name cast calculus terms and
% showed that the graduality proof could be modeled using semantics in
% certain kinds of \emph{double categories}, categories internal to the
% category of categories. A double category extends a category with a
% second notion of morphism, often a notion of ``relation'' to be paired
% with the notion of functional morphism, as well as a notion of
% functional morphisms preserving relations. In gradual typing the
% notion of relation models type precision and the squares model the
% term precision relation. This approach was influenced by the semantics
% of parametricity using reflexive graph categories
% \cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive graph
% categories are essentially double categories without a notion of
% relational composition. In addition to capturing the notions of type
% and term precision, the double categorical approach allows for a
% \emph{universal property} for casts: upcasts are the \emph{universal}
% way to turn a relation arrow into a function in a forward direction
% and downcasts are the dual universal arrow.

New and Licata \cite{new-licata18} developed an axiomatic account of the
graduality relation on a call-by-name cast calculus and showed that the
graduality proof could be modeled using semantics in certain kinds of double
categories.
%
Later, New, Licata and Ahmed \cite{new-licata-ahmed2019} extended this axiomatic
treatment from call-by-name to call-by-value as well by giving an axiomatic
theory of type and term precision in call-by-push-value. The double-categorical
framework serves as a foundation on which we base our model construction in this
paper. However, because of the lack of transitivity of the relation modeling
term precision, our model is not a double category because we cannot compose
squares horizontally.

%% With this notion of abstract categorical model in hand, denotational
%% semantics is then the work of constructing concrete models that
%% exhibit the categorical construction. New and Licata
%% \cite{new-licata18} present such a model using categories of
%% $\omega$-CPOs, and this model was extended by Lennon-Bertrand,
%% Maillard, Tabareau and Tanter to prove graduality of a gradual
%% dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
%% domain-theoretic approach meets our criteria of being a semantic
%% framework for proving graduality, but suffers from the limitations of
%% classical domain theory: the inability to model viciously
%% self-referential structures such as higher-order extensible state and
%% similar features such as runtime-extensible dynamic types. Since these
%% features are quite common in dynamically typed languages, we seek a
%% new denotational framework that can model these type system features.

%\subsubsection{Embedding-Projection Pairs}

A line of work by New, Licata and Ahmed proves graduality by
interpreting type precision $c : A \ltdyn A'$ as an
\emph{embedding-projection pair}, that is a pure function $e : A \to
A'$ and a possibly erroring projection $p : \li A' \multimap \li A$
such that $p \circ e = \id$ and $e \circ p \ltdyn \id$
\cite{new-ahmed2018,new-licata18,new-licata-ahmed2019}. At first glance, our model
of precision looks different: it is a \emph{relation}
$\sem{c} : A \rel A'$ with a quasi-representability structure that
gives us something like $e$ and $p$ in an embedding-projection
pair. The relationship between these models is that \emph{if} the
relation were \emph{truly} representable we would have that $e$ and $p$ form a \emph{Galois
connection} $e \circ p \ltdyn \id$ and $\id \ltdyn p \circ
e$. However, we have dropped the stronger property of retraction from
our analysis in this work. With true representability, the relation
$c$ is \emph{uniquely determined} by the embedding $e$, but since we
only have quasi-representability, we need to keep the relation around
explicitly. So the extra complexity of managing explicit relations is
a cost of the intensional reasoning that guarded type theory
introduces.

%\subsubsection{Step-Indexed Logical Relations Models}
The standard alternative to domain theory that scales to essentially
arbitrary self-referential definitions is \emph{step-indexing} or its
synthetic form of \emph{guarded recursion}. A series of works
\cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However,
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions.
%
% Additionally, the logical relations constructed to prove graduality in prior
% work \cite{new-ahmed2018,new-licata-ahmed2019,new-giovannini-licata-2022} suffer
% from technical complications of requiring two separate expression relations, one
% that counts steps on the left and the other on the right, and there is no
% analogue of this in our approach. However, using two expression relations allows
% some but not all transitive reasoning of term precision to be recovered. In the
% future we aim to explore if this approach is feasible in guarded semantics.
%
This is addressed somewhat by Siek and Chen
\cite{siek-chen2021}, who give a proof in Agda of graduality for an
operational semantics who use a \emph{guarded logic of propositions}
shallowly embedded in Agda. The guarded logic simplifies the treatment
of step-indexed logical relations, but the approach is still
fundamentally operational, and so the main lemmas of the work are
still tied to the particular operational syntactic calculus being
modeled. A further advantage of the denotational approach is that it
easily validates equational reasoning, not just graduality, and it is
completely independent of any particular syntax of gradual typing.

%\subsubsection{Gradual Dependent Types}
% Discuss Joey Eremondi's thesis on gradual dependent types
Eremondi \cite{Eremondi_2023} uses guarded type theory to
define a syntactic model for a gradually-typed source
language with dependent types. By working in guarded type theory, they are
able to give an exact semantics to non-terminating computations in a language
which is logically consistent, allowing for metatheoretic results about the
source language to be proven.
%
Similarly to our approach, they define a guarded lift monad to model potentially-
nonterminating computations and use guarded recursion to model the dynamic type.
However, they do not give a denotational semantics to term precision and it is unclear
how to prove graduality in this setting.
The work includes a formalization of the syntactic model in Guarded Cubical Agda.

% Discuss Jeremy Siek's work on graduality in Agda

\subsection{Mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper, we have
developed a partial formalization of our results in Guarded Cubical Agda
\cite{veltri-vezzosi2020}.
%
We formalized the major components of the definition of the concrete model
described in Section \ref{sec:concrete-relational-model}: predomains, error
domains, morphisms, relations, squares, using the guarded
features to define the free error domain, the lock-step error ordering, and weak
bisimilarity. We also formalized the no-go theorem from Section
\ref{sec:concrete-term-model}.

Building on these definitions, we formalized the notion of semantic
perturbations and push-pull structures as well as quasi-representable relations,
culminating in the definition of value and computation types and relations as
introduced in Section \ref{sec:concrete-relational-model}.
%
% culminating in the definition of value and computation types as predomains
% (resp. error domains) equipped with a monoid of syntactic perturbations with an
% interpretation homomorphism into the semantic perturbations. 
%
We constructed the predomain for the dynamic type via mixed induction and
guarded recursion, and defined its monoid of perturbations. We also defined the
relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
quasi-representable.
%
We have not yet formalized the constructions involving value and computation
relations: showing that these relations compose, and defining the actions of the
functors $\li$, $U$, $\times$, and $\arr$ on the relations. We have also not yet
formalized the syntax-to-semantics translation.

@@

@@

% \max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
In the future, we plan to apply our approach to give a denotational semantics
for languages that feature higher-order state or runtime-extensible dynamic
typing \cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
disciplines such as gradual dependent types and effect systems. This will
involve adapting prior operational models based on step-indexing (e.g.,
\cite{new-giovannini-licata-2022}) to the denotational setting and using guarded
type theory to obtain solutions to guarded domain equations as we did in this
work for the denotation of the dynamic type. The generality of our techniques
should allow for many of the constructions to be reused across different
languages. For instance, the free error domain construction can be easily
extended with additional cases to model effects besides error and stepping.
Additionally, we aim to complete our Agda formalization and evolve it into a
reusable framework for mechanized denotational semantics of gradually-typed
languages.

>>>>

@@

The methods of Abstracting Gradual Typing \cite{garcia-clark-tanter2016} and the
formal tools of the Gradualizer \cite{cimini-siek2016} provide a way to derive
from a statically-typed language a language that satisfies graduality by
construction. The main downside to these approaches lies in their inflexibility:
since the process in entirely mechanical, the language designer must adhere to
the predefined framework.  Many gradually typed languages do not fit into either
framework, e.g., Typed Racket \cite{tobin-hochstadt06, tobin-hochstadt08}, and
the semantics produced is not always the desired one.
%
Furthermore, while these frameworks do prove graduality of the resulting
languages, they do not show the correctness of the equational theory, which is
equally important to sound gradual typing.

% \subsubsection{Graduality via Embedding-Projection Pairs}

A line of work by New, Licata and Ahmed
\cite{new-ahmed2018,new-licata18,new-licata-ahmed2019} develops an axiomatic
account of gradual typing and proves graduality by interpreting type precision
$c : A \ltdyn A'$ as an \emph{embedding-projection pair}, that is, a pure
embedding function $e : A \to A'$ and a possibly erroring projection $p : \li A'
\multimap \li A$ such that $p \circ e = \id$ and $e \circ p \ltdyn \id$. At
first glance, our model of precision looks different: it is a \emph{relation}
$\sem{c} : A \rel A'$ with a quasi-representability structure that gives us
something close to $e$ and $p$ in an embedding-projection pair. The relationship
between these models is that \emph{if} the relation were \emph{truly}
representable we would have that $e$ and $p$ form a \emph{Galois connection} $e
\circ p \ltdyn \id$ and $\id \ltdyn p \circ e$. However, we have dropped the
stronger property of retraction from our analysis in this work. With true
representability, the relation $c$ is \emph{uniquely determined} by the
embedding $e$, but since we only have quasi-representability, we need to keep
the relation around explicitly. So the extra complexity of managing explicit
relations is a cost of the intensional reasoning that guarded type theory
introduces.

% Later, New, Licata and Ahmed \cite{new-licata-ahmed2019} extended this axiomatic
% treatment from call-by-name to call-by-value as well by giving an axiomatic
% theory of type and term precision in call-by-push-value.

% \subsubsection{Denotational and Operational Approaches}

The above axiomatic account of gradual typing is independent of the choice of
operational versus denotational semantics; we give an overview of the two
approaches below.

@@

On the denotational side, New and Licata \cite{new-licata18} showed that the
graduality proof could be modeled using semantics in certain kinds of double
categories, which are categories internal to the category of categories. A
double category extends a category with a second notion of morphism, often a
notion of ``relation'' to be paired with the notion of functional morphism, as
well as a notion of functional morphisms preserving relations. In gradual
typing, the notion of relation models type precision and the squares model the
term precision relation.
%
% This approach was influenced by the semantics of parametricity using reflexive
% graph categories \cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive
% graph categories are essentially double categories without a notion of
% relational composition. In addition to capturing the notions of type and term
% precision, the double categorical approach allows for a \emph{universal
% property} for casts: upcasts are the \emph{universal} way to turn a relation
% arrow into a function in a forward direction and downcasts are the dual
% universal arrow.
%
The double-categorical framework serves as a foundation on which we base our
model construction in this paper. However, our model is not actually a double
category: the lack of transitivity of bisimilarity means that we cannot compose
extensional squares horizontally.

With the notion of abstract categorical model for gradual typing in hand,
denotational semantics is then the work of constructing concrete models that
exhibit the categorical construction. New and Licata \cite{new-licata18} present
such a model using categories of $\omega$-CPOs, and this model was extended by
Lennon-Bertrand, Maillard, Tabareau and Tanter \cite{gradualizing-cic} to prove
graduality of a gradual dependently typed calculus $\textrm{CastCIC}^{\mathcal
G}$. This domain-theoretic approach meets our criterion of being a semantic
framework for proving graduality, but suffers from the limitations of classical
domain theory: the inability to model viciously self-referential structures such
as higher-order extensible state and similar features such as runtime-extensible
dynamic types.

% Since these features are quite common in dynamically typed languages, we have
% developed in this work a new denotational framework that can be extended to
% model these type system features.

%\subsubsection{Step-Indexed Logical Relations Models}
% The standard alternative to domain theory that scales to essentially
% arbitrary self-referential definitions is \emph{step-indexing} or its
% synthetic form of \emph{guarded recursion}.

On the operational side, a series of works
\cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
developed step-indexed logical relations models of gradually typed languages.
Unlike classical domain theory, such step-indexed techniques can scale to
essentially arbitrary self-referential definitions, which means they can model
higher-order store and runtime-extensible dynamic types
\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However, as is
common with operational approaches, their proof developments are highly
repetitive and technical, with each development formulating a logical relation
from first-principles and proving many of the same tedious lemmas without
reusable mathematical abstractions.
%
%
% Additionally, the logical relations constructed to prove graduality in prior
% work \cite{new-ahmed2018,new-licata-ahmed2019,new-giovannini-licata-2022} suffer
% from technical complications of requiring two separate expression relations, one
% that counts steps on the left and the other on the right, and there is no
% analogue of this in our approach. However, using two expression relations allows
% some but not all transitive reasoning of term precision to be recovered. In the
% future we aim to explore if this approach is feasible in guarded semantics.
%
This is addressed somewhat by Siek and Chen \cite{siek-chen2021}, who give a
proof in Agda of graduality for an operational semantics using a \emph{guarded
logic of propositions} shallowly embedded in Agda. The guarded logic simplifies
the treatment of step-indexed logical relations, but the approach is still
fundamentally operational, and so the main lemmas of the work are still tied to
the particular operational syntactic calculus being modeled.
% A further advantage of the denotational approach is that it easily validates
% equational reasoning, not just graduality, and it is completely independent of
% any particular syntax of gradual typing.

Our work combines the benefits of both the denotational and step-indexed
approaches. We take as a starting point the denotational approach of New and
Licata, but we work with guarded type theory (a synthetic form of step-indexing)
rather than classical domain theory. In the guarded setting, the compositional
theory of New-Licata breaks down: when accounting for steps, the casts are no
longer embedding-projection pairs, and we cannot carry out compositional
reasoning on terms when those terms differ by an arbitrary unknown number of
steps. Our work provides a novel refinement to their theory that allows for some
compositional reasoning to be recovered even in the guarded setting.

% \subsubsection{Gradual Dependent Types}
Eremondi \cite{Eremondi_2023} uses guarded type theory to
define a syntactic model for a gradually-typed source
language with dependent types. By working in guarded type theory, they are
able to give an exact semantics to non-terminating computations in a language
which is logically consistent, allowing for metatheoretic results about the
source language to be proven.
%
Similarly to our approach, they define a guarded lift monad to model potentially-
nonterminating computations and use guarded recursion to model the dynamic type.
However, they do not give a denotational semantics to term precision and it is unclear
how to prove graduality in this setting.
Their work includes a formalization of the syntactic model in Guarded Cubical Agda.

\subsection{Comparison of Denotational and Operational Approaches}
\eric{Some of this could be folded into the previous section, but this was one of the
big-picture questions from one of the reviewers.}

Denotational methods have several benefits compared to operational approaches,
and this carries over to the setting of gradual typing. First, denotational
methods are compositional and reusable, and allow for the use of existing
mathematical constructs and theorems, e.g., partial orderings, monads, etc,
while operational methods tend to require a significant amount of boilerplate
work to be done from scratch in each new development.
%
As a specific example of the compositional nature of our approach, the treatment
of the cast rules in our work is more compositional than in previous work using
operational semantics. Recall from Section \ref{sec:towards-relational-model}
that the cast rules needed for the proof of graduality build in composition of
type precision derivations. Rather than proving these from scratch in the model,
we are able to take as primitive simpler versions of the cast rules whose
validity in the model is easier to establish. Then from these simpler rules, we
derive the original ones using compositional reasoning.

A further advantage of the denotational approach is that it is completely
independent of any particular syntax of gradual typing. This allows for reuse
across multiple languages and makes it particularly straightforward to
accommodate additions to a language: adding support for a new type amounts to
defining a new object and extending the dynamic type accordingly. It also may be
possible to model alternative cast semantics by changing the definition of the
dynamic type and some type casts.
%
In contrast, operational methods are not as readily extensible, generally
requiring adding cases to the logical relations and the inductive proofs. In a
mechanized metatheory, this manifests as a ``copy-paste'' reusability rather than
the true code reuse one obtains with a denotational semantics.
%
An additional benefit to the denotational approach is that it is trivial to
establish the validity of the $\beta$ and $\eta$ principles, because they are
equalities in the semantics, whereas they require tedious manual proof in the
operational approach.

Finally, while we advocate for a denotational approach, the overall structure of
our semantics could be applied to an operational logical relations proof. For
example, there could be a step-indexed logical relation for strong error
ordering and one for weak bisimilarity, corresponding to the fact that objects
in our denotational model have an error ordering and a bisimilarity relation.
Then the precision rules for casts could be similarly validated using
perturbations as we did in this work.

\subsection{Implications for Gradual Typing Research}

While we have focused on a single semantics in this work, we expect that our
development can be adapted to prove graduality and (more limited) equational
reasoning principles for other cast semantics.
%
In addition, while we have focused on proving extensional properties in this
work, the intensional nature of our semantics can be viewed as a form of
\emph{cost semantics}, where a runtime cost is incurred from inspecting the
dynamic type. This could be used to verify that cast optimizations such as space
efficient implementations \cite{herman-tomb-flanagan-2010} are not only
extensionally correct, but also have lower quantitative cost. We leave this as
future work.

\subsection{Mechanization}\label{sec:mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper, we have
developed a partial formalization of our results in Guarded Cubical Agda
\cite{veltri-vezzosi2020}.
%
We formalized the major components of the definition of the concrete model
described in Section \ref{sec:concrete-relational-model}: predomains, error
domains, morphisms, relations, squares, using the guarded
features to define the free error domain, the lock-step error ordering, and weak
bisimilarity. We also formalized the no-go theorem from Section
\ref{sec:concrete-term-model}.

Building on these definitions, we formalized the notion of semantic
perturbations and push-pull structures as well as quasi-representable relations,
culminating in the definition of value and computation types and relations as
introduced in Section \ref{sec:concrete-relational-model}.
%
% culminating in the definition of value and computation types as predomains
% (resp. error domains) equipped with a monoid of syntactic perturbations with an
% interpretation homomorphism into the semantic perturbations. 
%
We constructed the predomain for the dynamic type via mixed induction and
guarded recursion, and defined its monoid of perturbations. We also defined the
relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
quasi-representable.

% We have not yet formalized the constructions involving value and computation
% relations: showing that these relations compose, and defining the actions of the
% functors $\li$, $U$, $\times$, and $\arr$ on the relations. We have also not yet
% formalized the syntax-to-semantics translation.

@@

The remaining formalization work is the following:
\begin{enumerate}
    \item Showing that the functors $\times$ and $\arr$ preserve
    quasi-representability, which requires tedious reasoning about the Kleisli actions of
    $\times$ and $\arr$. These results are proved in the Appendix.
    \item Verifying the rules in the model corresponding to the equations for
    type precision derivations (see the bottom of Figure \ref{fig:gtlc-syntax}).
    \item Defining the syntax-to-semantics translation.
\end{enumerate}

@@

% \max{Expand on this, e.g. what some of the challenges might be and what would be reusable}
In the future, we plan to apply our approach to give a denotational semantics
to gradually-typed languages that feature higher-order state or runtime-extensible dynamic
typing \cite{DBLP:journals/corr/abs-2210-02169} as well as richer type
disciplines such as gradual dependent types and effect systems. This will
involve adapting prior operational models based on step-indexing (e.g.,
\cite{new-giovannini-licata-2022}) to the denotational setting and using guarded
type theory to obtain solutions to guarded domain equations as we did in this
work for the denotation of the dynamic type. The generality of our techniques
should allow for many of the constructions to be reused across different
languages. For instance, the free error domain construction can be easily
extended with additional cases to model effects besides error and stepping.
Additionally, we aim to complete our Agda formalization and evolve it into a
reusable framework for mechanized denotational semantics of gradually-typed
languages.
